---
title: "CMIP5 Interactive vizualizations"
output: html_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = F, cache = F, message = F, warning = F)
```

```{r packages}

library(hydrosystems); 
library(ggHydro); 
library(lubridate); 
library(readr); 
library(readxl);
library(dplyr);
library(tidyr); 
library(highcharter)
```

Read GCM data from the CMIP5 data directory

```{r readData}
gcm_data <- read_rds("gcm_data.Rds")
```

## User defined variables

Control the plots through these objects:

- **data** R object that stores monthly CMIP5 projections for all model/scenario combinations 

- **hist.period** historical climate period (e.g., 1905:2005)

- **proj.period** projected climate period (e.g., 2006:2095)

- **variable** variable to be plotted (currently "tavg" or "precip")

<br>

## CMIP5 Scatter Plot

Use highcharter tools for interactive scatter of GCM projections

```{r cmip5Scatter-source}

cmip5ScatterPlotInteractive <- function(
  data = NULL, hist.period = NULL, proj.period = NULL,
  tavg.breaks = NULL, prcp.breaks = NULL)
{

  rcp_col <- c("#08519c", "#abd9e9", "#fe9929", "#f03b20")
 
  # Summarize data for each projection
  df <- lapply(names(data),
               function(x) bind_rows(data[[x]], .id = "model")) %>%
    setNames(names(data)) %>%
    bind_rows(.id = "scenario") %>% ungroup() %>%
    mutate(scenario = factor(scenario, 
      levels = c("historical", "rcp26", "rcp45","rcp60","rcp85"),
      labels = c("Historical", "RCP2.6", "RCP4.5", "RCP6.0", "RCP8.5")))
  
  data_hist <- df %>% filter(scenario == "Historical") %>%
    filter(year %in% hist.period) %>%
    group_by(model) %>%
    summarize_at(vars(prcp, tavg), mean) %>%
    mutate(prcp = prcp * 12)
  
  data_proj <- df %>% filter(scenario != "Historical") %>%
    filter(year %in% proj.period) %>%
    group_by(scenario, model) %>%
    summarize_at(vars(prcp, tavg), mean) %>%
    mutate(prcp = prcp * 12)
  
  # mean precip and temp changes
  delta_prcp <- data_proj %>%
    select(scenario, model, prcp) %>%
    left_join(select(data_hist, model, hist_prcp = prcp), by = "model") %>%
    mutate(prcp = round((prcp - hist_prcp) / hist_prcp * 100,2))
  
  delta_tavg <- data_proj %>%
    select(scenario, model, tavg) %>%
    left_join(select(data_hist, model, hist_tavg = tavg), by = "model") %>%
    mutate(tavg = round(tavg - hist_tavg,2))
  
  delta_clim <- delta_prcp %>%
    left_join(delta_tavg, by = c("scenario", "model")) %>%
    na.omit() %>% select(scenario, model, prcp, tavg)
  
  # Axis breaks (if not provided)
  if(is.null(tavg.breaks)) {
    tavg.breaks <- seq(0, round(max(delta_tavg$tavg, na.rm = T),0) + 2, 1)
  }
  if(is.null(prcp.breaks)) {
    prcp.breaks <- seq(
      round(min(delta_prcp$prcp, na.rm = TRUE),-1) -20,
      round(max(delta_prcp$prcp, na.rm = TRUE),-1) +20,
      10)
  }
  
  # Axis limits
  tavg_step <- (tavg.breaks[2] - tavg.breaks[1])/2
  tavg_lim  <- range(tavg.breaks) + c(- tavg_step, tavg_step)
  prcp_step <- (prcp.breaks[2] - prcp.breaks[1])/2
  prcp_lim  <- range(prcp.breaks) + c(- prcp_step, prcp_step)
  
  highchart(width = 850, height = 750) %>% 
    hc_add_series_df(delta_clim, 
                     type="scatter", radius=7,
                     x=tavg, y=prcp, group=scenario, allowPointSelect = T,
                     borderwidth = 2) %>%
    #Axis settings
    hc_xAxis(min = min(tavg.breaks), max = max(tavg.breaks), 
             tickInterval= tavg.breaks[2] - tavg.breaks[1], 
             gridLineWidth=2, crosshair=T,
             title = list(text = "Temperature change (°C)")) %>%
    hc_yAxis(min = min(prcp.breaks), max = max(prcp.breaks), 
             tickInterval= prcp.breaks[2] - prcp.breaks[1],
    gridLineWidth=2, crosshair=T,
             title = list(text="Precipitation change (%)")) %>%
    #Plot appearance
    hc_add_theme(hc_theme_smpl()) %>%
    hc_plotOptions(
      series = list(marker = list(symbol = "circle"))) %>%
    hc_legend(align = "right", verticalAlign = "top", 
              layout = "vertical", x = 0, y = 50) %>%
    hc_colors(rcp_col) %>%
    #Additional information & settings
    hc_tooltip(formatter = JS("function(){
        return (' Scenario: ' + this.point.scenario +
                ' <br> Model :' + this.point.model +
                ' <br> ΔTemp (°C) :' + this.x +
                ' <br> ΔPrecip (%) :' + this.y)}"),
      borderWidth = 2) %>%
    hc_exporting(enabled = T)
}

```

```{r cmip5Scatter, echo = TRUE}
cmip5ScatterPlotInteractive(
  data = gcm_data,
  hist.period = 1976:2005, 
  proj.period = 2066:2095, 
  tavg.breaks = seq(0,8,1), 
  prcp.breaks = seq(-40,40,10))
```

<br>

## CMIP5 IPCC-Style Trends 

Use highcharter tools for IPCC style GCM trends

```{r cmip5Trends-source}

# Source scripts
cmip5TrendsInteractive <- function(data = NULL, variable = NULL, 
  hist.period = NULL, proj.period = NULL, ref.period = 1986:2005)
{

  rcp_col <- c("black", "#08519c", "#abd9e9", "#fe9929", "#f03b20")
  
  hist_ref <- data[["historical"]] %>% bind_rows(.id = "model") %>%
    filter(year %in% ref.period) %>% ungroup() %>%
    summarize(prcp = mean(prcp) * 12, tavg = mean(tavg))
  
  hist <- data[["historical"]] %>% bind_rows(.id = "model") %>% 
    mutate(scenario = "Historical") %>% filter(year %in% hist.period)
  
  rcp1 <- data[["rcp26"]] %>% bind_rows(.id = "model") %>% mutate(scenario = "RCP2.6")
  rcp2 <- data[["rcp45"]] %>% bind_rows(.id = "model") %>% mutate(scenario = "RCP4.5")
  rcp3 <- data[["rcp60"]] %>% bind_rows(.id = "model") %>% mutate(scenario = "RCP6.0")
  rcp4 <- data[["rcp85"]] %>% bind_rows(.id = "model") %>% mutate(scenario = "RCP8.5")
  rcp  <- bind_rows(rcp1, rcp2, rcp3, rcp4) %>% filter(year %in% proj.period)
  
  clim_abs <- bind_rows(hist, rcp) %>%
    group_by(model, scenario, year) %>%
    summarize(prcp = mean(prcp)*12, tavg = mean(tavg))
  
  clim_anm <- clim_abs %>% ungroup() %>%
    mutate(prcp = prcp/hist_ref$prcp * 100 - 100) %>%  #relative change 
    mutate(tavg = tavg - hist_ref$tavg) # increase
  
  df <- clim_anm %>%
    select(model:year, var = variable) %>%
    group_by(scenario, year) %>%
    summarize(mean = mean(var), min = quantile(var, probs = 0.05), 
              max = quantile(var, probs = 0.95)) %>% ungroup()
  
  
  df1 <- df %>% select(scenario, x = year, low = min, high = max) %>%
    mutate(low = round(low,2), high = round(high,2)) %>%
    mutate(scenario = factor(scenario, 
      levels = c("Historical", "RCP2.6", "RCP4.5", "RCP6.0", "RCP8.5"),
      labels = c("Historical-Range", "RCP2.6-Range", "RCP4.5-Range", "RCP6.0-Range", "RCP8.5-Range")))
  
  df2 <- df %>% select(scenario, x = year, y = mean) %>%
    mutate(y = round(y, 2)) %>% 
    mutate(scenario = factor(scenario, 
      levels = c("Historical", "RCP2.6", "RCP4.5", "RCP6.0", "RCP8.5"),
      labels = c("Historical-Mean", "RCP2.6-Mean", "RCP4.5-Mean", "RCP6.0-Mean", "RCP8.5-Mean")))
  
  highchart() %>%
    hc_chart(borderWidth = 0.1) %>%
    hc_add_theme(hc_theme_smpl()) %>%
    hc_add_series(df1, type = "arearange", lineWidth = 0.1,
                  hcaes(color = "scenario", group = "scenario"),
                  fillOpacity = 0.2, showInLegend = T) %>%
    hc_add_series(df2, type = "line",
                  hcaes(color = "scenario", group = "scenario")) %>%
    hc_colors(rcp_col) %>%
    hc_legend(align = "right", verticalAlign = "top", 
              layout = "vertical", x = 0, y = 50) %>%
    hc_tooltip(crosshairs = TRUE,
             headerFormat = "This is a custom header<br>",
             formatter = JS("function(){ return (' Scenario: ' 
                  + this.point.scenario +' <br> Year: ' + this.x + ' <br> Value: ' + this.y)}"),
             borderWidth = 2, borderColor = "black") %>%
    hc_exporting(enabled = T) %>%
    hc_xAxis(title = list(text = "Year"),
           plotLines = list(
             list(color = "#e0e0e0", width = 2, value = 1900),
             list(color = "#e0e0e0", width = 2, value = 1950),
             list(color = "#e0e0e0", width = 2, value = 2000),
             list(color = "#e0e0e0", width = 2, value = 2050),
             list(color = "#e0e0e0", width = 2, value = 2100))) %>%
    hc_yAxis(tickPositions = c(-6, -4, -2, 0, 2, 4, 6, 8, 10, 12)) %>%
      hc_plotOptions(
     marker = list(
       lineWidth = 1,
       fillColor = "black",
       lineColor = "#ffffff"
       )
     )  
}
```

```{r cmip5Trends, echo = TRUE}
cmip5TrendsInteractive(
  data = gcm_data, 
  variable = "tavg", 
  hist.period = 1890:2005, 
  proj.period = 2006:2100)
```

<br>

## CMIP5 Seasonal Changes (Boxplot)

```{r cmip5Seasonal-source, echo = FALSE}

cmip5SeasonalInteractive <- function(data = NULL, variable = NULL, 
  hist.period = NULL, proj.period = NULL)
  {
  
  rcp_col <- c("#08519c", "#abd9e9", "#fe9929", "#f03b20")
  
  # Summarize data for each projection
  df <- lapply(names(data),
               function(x) bind_rows(data[[x]], .id = "model")) %>%
    setNames(names(data)) %>%
    bind_rows(.id = "scenario") %>% ungroup() %>%
    mutate(scenario = factor(scenario, 
                             levels = c("historical", "rcp26", "rcp45","rcp60","rcp85"),
                             labels = c("Historical", "RCP2.6", "RCP4.5", "RCP6.0", "RCP8.5")))
  
  data_hist <- df %>% filter(scenario == "Historical") %>%
    filter(year %in% hist.period) %>%
    group_by(model, mon) %>%
    summarize_at(vars(prcp, tavg), mean) %>%
    mutate(prcp = prcp * 12)
  
  data_proj <- df %>% filter(scenario != "Historical") %>%
    filter(year %in% proj.period) %>%
    group_by(scenario, model, mon) %>%
    summarize_at(vars(prcp, tavg), mean) %>%
    mutate(prcp = prcp * 12)
  
  # mean precip and temp changes
  delta_prcp <- data_proj %>%
    select(scenario, model, prcp, mon) %>%
    left_join(select(data_hist, model, hist_prcp = prcp, mon), by = c("model", "mon")) %>%
    mutate(prcp = round((prcp - hist_prcp) / hist_prcp * 100,2))
  
  delta_tavg <- data_proj %>%
    select(scenario, model, tavg, mon) %>%
    left_join(select(data_hist, model, hist_tavg = tavg, mon), by = c("model", "mon")) %>%
    mutate(tavg = round(tavg - hist_tavg,2))
  
  delta_clim <- delta_prcp %>%
    left_join(delta_tavg, by = c("scenario", "model", "mon")) %>%
    na.omit() %>% select(scenario, model, mon, prcp, tavg)
  
  df <- delta_clim %>% select(scenario, model, mon, var = variable) %>%
    mutate(mon = factor(mon, levels = 1:12, labels = month.abb))

  # Axis breaks (if not provided)
  yaxis_title <- if(variable == "tavg") {"Temperature change (°C)"} else {"Precip. change (%)"}
  
  hcboxplot(x = df$var, var = df$mon, var2 = df$scenario, outliers = FALSE) %>%
    hc_chart(type = "column") %>%
    hc_colors(rcp_col) %>%
    hc_legend(align = "right", verticalAlign = "top", 
              layout = "vertical", x = 0, y = 50) %>%
    hc_tooltip(crosshairs = TRUE, borderWidth = 2, borderColor = "black") %>%
    hc_exporting(enabled = T) %>%
    hc_yAxis(title = list(text = yaxis_title)) %>%
    hc_xAxis(title = list(text = "Month")) 
}

```

```{r cmip5Seasonal, echo = TRUE}

cmip5SeasonalInteractive(
  data = gcm_data, 
  variable = "tavg",
  hist.period = 1890:2005,
  proj.period = 2006:2100)

```
